Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response

### 2024-12-07 paper/paper.qmd
update parquet data in use and move bias/privacy of data into dicussion section

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(knitr)

fire_data <- read.csv(file = here::here("data/02-analysis_data/tfs_analysis_data"))

@@

## Data Cleaning

After loading the raw data, I selected the columns of interest as explained in the previous section. These columns were then renamed to lowercase, and spaces were replaced with underscores to ensure consistency. After reviewing the data, I kept in mind that this study focuses on the relationship between ignition sources and fire severity. As a result, I removed rows with "Undetermined" ignition sources.

@@

```{r}
#| label: fig-response-time-distribution
#| fig-cap: Distribution of response times for fire incidents in Toronto, showing the frequency of various response times (in minutes). This visualization highlights the efficiency of the Toronto Fire Service's (TFS) response. It's almost normally distributed with a mean of around 5, and the spread of not too far.
#| echo: false
#| warning: false
library(ggplot2)
fire_data <- read.csv(here::here("data/02-analysis_data/tfs_analysis_data"))
## 15. Response Time (Numerical) ##
#Plot
ggplot(fire_data, aes(x = response_time)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  theme_minimal() +
  ggtitle("Response Time Distribution") +
  xlab("Response Time (minutes)") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = 0.5, size = 3) +
  xlim(0, 15)

@@

  # Public and Performance Spaces
  else if (grepl("Stage|Rink|Gym|Performance|Concourse|Court", area)) {
    return("Public and Performance Spaces")
  } 
  # Miscellaneous and Unclassified
  else {
    return("Miscellaneous and Unclassified")
  }
}
# Apply the categorization function to each area of origin
fire_data$area_of_origin_grouped <- sapply(fire_data$area_of_origin, 
                                           categorize_area_of_origin)
# Plot
ggplot(fire_data, aes(x = area_of_origin_grouped)) +
  geom_bar(fill = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Area of Origin") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

@@

ggplot(fire_data, aes(x = final_incident_type)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  ) +
  scale_x_discrete(labels = function(x) sapply(x, function(y) paste(head(strsplit(y, " ")[[1]], 3), collapse = " "))) +
  xlab("Final Incident Type") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

@@

  # Vehicle and Accidents
  else if (grepl("Vehicle|Accident|Collision", source)) {
    return("Vehicle and Accidents")
  } 
  # Multiple Ignition Sources (or suspected arson)
  else if (grepl("Multiple|Arson", source)) {
    return("Multiple Ignition Sources")
  } 
  # Cooling and Air Conditioning Equipment
  else if (grepl("Cooling|Air Conditioner", source)) {
    return("Cooling and Air Conditioning Equipment")
  } 
  # Other (for anything that doesn't fit into the above categories)
  else {
    return("Other")
  }
}
# Apply the categorization function to each ignition source
fire_data$ignition_source_grouped <- sapply(fire_data$ignition_source, categorize_ignition_sources)
#Table
#Plot
# Plot of grouped ignition source distribution
ggplot(fire_data, aes(x = ignition_source_grouped)) +
  geom_bar(fill = "green") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("Ignition Source Group") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5)

@@

fire_data_model <- read_csv( file = here::here("data/02-analysis_data/severity_tfs_analysis_data"))

@@

#### Read data ####
fire_data_model <- read_csv(file = here::here("data/02-analysis_data/updated_tfs_analysis_data"))

@@

fire_data_model <- read_csv(file = here::here("data/02-analysis_data/severity_tfs_analysis_data"))

@@

## Fire Alarm System Status

@@

fire_data_current <- read_csv( file = here::here("data/02-analysis_data/severity_tfs_analysis_data"))

@@

@@

## Government and Agency Actions for Improving Fire Safety Infrastructure {#sec-second-point}

@@

## TFS Decision Making {#sec-third-point}

@@

fire_data_model <- read_csv(here::here("data/02-analysis_data/severity_tfs_analysis_data"))

@@

# Appendix A {#sec-appendixa}

@@

#### Read data ####
fire_data <- read.csv(here::here("data/02-analysis_data/tfs_analysis_data"))
#### Data Overview ####
# Structure
head(fire_data)
str(fire_data)

@@

This appendix contains all plot/summary distribution of all variables in intrested choose in analysis data set.

@@

# Appendix B {#sec-appendix-B}

## Method Behind Data

### Potential Bias

The fire incidence dataset includes several categorical variables such as Ignition Source, Possible Cause, and Initial Types. During data collection, these variables, which have a wide range of categories, may introduce various types of bias.

### **Privacy Protect**

### Sampling Method in Fire Incident Data

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(knitr)
library(arrow)

fire_data <- read_parquet(file = here::here("data/02-analysis_data/tfs_analysis_data.parquet"))

@@

## Data Cleaning
The raw dataset was stored in its original csv format to ensure data integrity and reproducibility, and later analysis data after cleaning and processing is saved into parquet file using `write_parquet` function in `arrow` package [@arrow2024].

After loading the original raw data, I selected the columns of interest as explained in the previous section. These columns were then renamed to lowercase, and spaces were replaced with underscores to ensure consistency. After reviewing the data, I kept in mind that this study focuses on the relationship between ignition sources and fire severity. As a result, I removed rows with "Undetermined" ignition sources.

@@

```{r}
#| label: fig-response-time-distribution
#| fig-cap: Distribution of response times for fire incidents in Toronto, showing the frequency of various response times (in minutes). This visualization highlights the efficiency of the Toronto Fire Service's (TFS) response. It's almost normally distributed with a mean of around 5, and the spread of not too far.
#| echo: false
#| warning: false
library(ggplot2)
library(arrow)
fire_data <- read_parquet(here::here("data/02-analysis_data/tfs_analysis_data.parquet"))
## 15. Response Time (Numerical) ##
#Plot
ggplot(fire_data, aes(x = response_time)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  theme_minimal() +
  ggtitle("Response Time Distribution") +
  xlab("Response Time (minutes)") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = -0.5, size = 3) +
  xlim(0, 15)

@@

  # Public and Performance Spaces
  else if (grepl("Stage|Rink|Gym|Performance|Concourse|Court", area)) {
    return("Public and Performance Spaces")
  } 
  # Miscellaneous and Unclassified
  else {
    return("Miscellaneous and Unclassified")
  }
}
# Apply the categorization function to each area of origin
fire_data$area_of_origin_grouped <- sapply(fire_data$area_of_origin, 
                                           categorize_area_of_origin)
# Plot
ggplot(fire_data, aes(x = area_of_origin_grouped)) +
  geom_bar(fill = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Area of Origin") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)

@@

ggplot(fire_data, aes(x = final_incident_type)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  ) +
  scale_x_discrete(labels = function(x) sapply(x, function(y) paste(head(strsplit(y, " ")[[1]], 3), collapse = " "))) +
  xlab("Final Incident Type") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)

@@

  # Vehicle and Accidents
  else if (grepl("Vehicle|Accident|Collision", source)) {
    return("Vehicle and Accidents")
  } 
  # Multiple Ignition Sources (or suspected arson)
  else if (grepl("Multiple|Arson", source)) {
    return("Multiple Ignition Sources")
  } 
  # Cooling and Air Conditioning Equipment
  else if (grepl("Cooling|Air Conditioner", source)) {
    return("Cooling and Air Conditioning Equipment")
  } 
  # Other (for anything that doesn't fit into the above categories)
  else {
    return("Other")
  }
}
# Apply the categorization function to each ignition source
fire_data$ignition_source_grouped <- sapply(fire_data$ignition_source, categorize_ignition_sources)
#Table
#Plot
# Plot of grouped ignition source distribution
ggplot(fire_data, aes(x = ignition_source_grouped)) +
  geom_bar(fill = "green") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("Ignition Source Group") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)

@@

fire_data_model <- read_parquet( file = here::here("data/02-analysis_data/severity_tfs_analysis_data.parquet"))

@@

#### Read data ####
fire_data_model <- read_parquet(file = here::here("data/02-analysis_data/updated_tfs_analysis_data.parquet"))

@@

fire_data_model <- read_parquet(file = here::here("data/02-analysis_data/severity_tfs_analysis_data.parquet"))

@@

### Fire Alarm System Status

@@

fire_data_current <- read_parquet( file = here::here("data/02-analysis_data/severity_tfs_analysis_data.parquet"))

@@

### Sprinkler System Status

@@

## Limitations and Biases in the Data {#sec-second-point}
### Potential Bias

The fire incidence dataset includes several categorical variables such as Ignition Source, Possible Cause, and Initial Types. During data collection, these variables, which have a limited range of categories, may introduce various types of bias.

### Sampling Method in the Data

### Privacy Protection in the Data

## Government and Agency Actions for Improving Fire Safety Infrastructure {#sec-third-point}

@@

## TFS Decision Making {#sec-fourth-point}

@@

fire_data_model <- read_parquet(here::here("data/02-analysis_data/severity_tfs_analysis_data.parquet"))

@@

# Appendix {-}

# Data and Variable Selection

@@

#### Read data ####
fire_data <- read_parquet(here::here("data/02-analysis_data/tfs_analysis_data.parquet"))
#### Data Overview ####
# Structure
head(fire_data)
str(fire_data)

@@

This appendix contains all plot/summary distribution of all variables in interested choose in analysis data set.

@@
